\begin{abstract}
    This work will discuss \textbf{computational approaches} developed to refine and improve \textit{current cancer treatments}.

    \textbf{Cancer} is one of the \textit{leading causes of death} globally, responsible for millions of deaths each year. It is a complex \textit{group of diseases} characterized by the \textit{uncontrolled growth and spread of abnormal cells} in the body, driven by a multitude of \textbf{genetic mutations}. These cells can invade surrounding tissues and, if left untreated, may spread to other parts of the body. Therefore, it is essential to search for an \textit{effective treatment}, which not only alleviates symptoms and improves the quality of life, but also addresses the underlying mechanisms of cancer.

    Over the years, \textit{various strategies} have been developed and implemented to treat cancer as effectively as possible. However, most currently employed cancer treatments have significant \textbf{side effects}, which can severely affect patients' quality of life. As a result, there is a pressing need for therapies that are not only effective in \textit{targeting cancer cells}, but also \textit{minimize harm to healthy tissues}, thereby improving the patient's overall experience during treatment.

    This work will focus specifically on one type of treatment: \textbf{targeted therapies}, which have shown significant results in recent years, both in terms of \textit{reducing side effects} and \textit{improving treatment efficacy}. Targeted therapies, as the name suggests, are designed to focus on specific molecules or pathways involved in cancer growth and progression. By precisely \textbf{targeting} these elements, targeted therapies aim to disrupt the cancer's ability to proliferate, while \textit{minimizing damage} to healthy cells, resulting in improved treatment outcomes and fewer side effects.

    However, \textbf{identifying effective targets} is not straightforward, as it depends on a variety of factors. In fact, throughout the years, numerous studies have attempted to differentiate between the \textit{vast number of mutations} that occur in cancer, aiming to identify \textit{the most relevant} ones for tumor treatment.

    Fortunately, certain \textbf{phenomena in genomic data} have been identified that could potentially be leveraged to simplify the search in this field. This work will explore the \textbf{computational approaches} adopted in recent years, the challenges this search presents, and the \textit{genomic characteristics} that newly developed algorithms leverage to \textbf{classify mutations efficiently}.

    The first chapter of this work will provide an \textbf{overview of cancer}, exploring its causes, current treatment options, and a more detailed examination of \textit{targeted therapies}. The second chapter will highlight the challenges associated with \textbf{differentiating} among the vast number of mutations involved in cancer and will \textit{introduce the studies} that will be analyzed throughout this work, focusing specifically on how they \textbf{mathematically formulated} biological phenomena observed \textit{statistically} in genomic data. The third chapter will delve deeper into the \textbf{algorithms} employed by the discussed studies, and explain how they employed their own \textit{metrics}. Finally, the last chapter will address \textbf{additional considerations} regarding the studies presented.
\end{abstract}

\let\cleardoublepage\clearpage
