\chapter*{Acknowledgements}

\addcontentsline{toc}{chapter}{Acknowledgements}

This work marks the conclusion of a journey that began three years ago, a path I embarked on with dedication and commitment. 

First and foremost, I would like to express my deepest gratitude to my advisor, who closely supported me throughout the entire process of research and thesis writing. His regular feedback and suggestions were invaluable and significantly contributed to the quality of this work. Without his consistent guidance and help, this thesis would not have reached the level it has achieved.

A heartfelt thanks goes to my parents, who have stood by me throughout my academic journey, not just during university, but long before. I know I can continue to count on their unwavering support as I move forward with my future endeavors. I am also deeply grateful to my entire family, who have always believed in my life choices and have consistently supported me in achieving my goals to the best of my ability.

I extend my sincere thanks to my university colleagues, with whom I have formed strong bonds that go far beyond the academic setting. Sharing this university experience with them has allowed me to grow both intellectually and emotionally, making this journey unforgettable.

I would also like to acknowledge my friends from earlier stages of life, whose bonds may have weakened over time but who laid the foundation for the person I am proud to be today.

Finally, my gratitude goes out to everyone who has helped me reach this point, including those who may not have been mentioned directly. Your impact has been deeply felt, and I am truly thankful.

